Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?

### 2023-10-18 ms/ms.Rmd
revised discussion

---
# This template is largely functionally equivalent to that for elsevier
# The two fields that are not in the elsevier format are:
#      `corresponding_author` and `acknowledgements`
title: Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous *Arabidopsis thaliana* Incurs No Photosynthetic Advantage
#date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: jwatts@colgate.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: Colgate University
  - name: Graham Dow
    email: graham.dow@usys.ethz.ch
    affiliation: ETH
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: University of California Davis
    footnote: 1 # Footnote using reference
  - name: Christopher D. Muir
    email: cdmuir@hawaii.edu
    affiliation: University of Hawaii at Manoa
    footnote: 1 # Footnote using reference
address:
  - code: Colgate University
    address: 13 Oak Drive, Hamilton, NY 13346
  - code: University of Hawaii at Manoa
    address: 2500 Campus Rd, Honolulu, HI 96822
footnote:
  - code: 1
    text: "Equal contribution"
  - code: 2
    text: "Current email address: \\href{mailto:cat@example.com}{cat@example.com}"
abstract: |
  This is the abstract.
  It consists of two paragraphs.
acknowledgements: |
  This is an acknowledgement.
  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE) 
knitr::opts_chunk$set(fig.pos = 'p') # Places figures on pages separate from text
knitr::opts_chunk$set(out.width = '100%', dpi = 300) # Figure resolution and size
knitr::opts_chunk$set(fig.env = "figure") # Latex figure environment
knitr::opts_knit$set(root.dir = "..")
```

@@

Many successful predictions about stomata and other leaf traits can be made by hypothesizing that natural selection should optimizes CO$_2$ gain per unit of water loss [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017]. However, stomatal anatomy may be partially constrained by physical and developments limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023]. Sometimes optimization leads to similar phenotypes across many disparate species. For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning [@geisler_oriented_2000; @dow_physiological_2014]; however some species (notably in *Begonia*) appear to benefit from overlapping vapor shells caused by stomatal clustering [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017]. Stomatal traits also vary adaptively in different environments. Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions. Stomatal size is jointly controlled by genome size, light, and stomatal density [@jordan_environmental_2015]. Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011]. Total stomatal area (size $\times$ density) is optimized for operational conductance ($gs_\text{s,op}$) rather than maximum conductance ($gs_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021]. Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $gs_\text{s,max}$ [@sack_developmental_2016], which is proportional to $gs_\text{s,op}$ [@murray_consistent_2020]. Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020]. In most species, stomata occur on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019]. Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991]. Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are poorly understood and likely more complex than random development along the leaf surface. @croxdale_stomatal_2000] reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.

@@

We hypothesized that natural selection will favor stomatal patterning and distribution to minimize the diffusion path length. In amphistomatous leaves, this would be accomplished by 1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface. Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light. Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest. We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$). This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
\label{fig:ideal-amphi-grid}
\end{figure}

@@

df = tibble::tribble(
  ~trait, ~relationship,
  "leaf thickness", "+",
  "stomatal density", "-",
  "leaf porosity", "-", 
  "lat.-vert. diffusion ratio", "-", 
  "temperature", "-",
  "pressure", "-",
  "Rubisco concentration", "+",
  "light", "+"
)

@@

[CDM: Graham Dow provided these images. We'll need to add him as a co-author and ask him to write methods on image acquisition.]

@@

For each sample image, we also simulated $10^3$ synthetic data with $n$ stomata ideally dispersed in an equilateral triangular grid. For these grids, we integrated over plausible stomatal densities and then conditioned on stomatal grids with exactly $n$ stomata. The simulated stomatal count was drawn from a Poisson distribution with the mean parameter $\lambda$ drawn from a Gamma distribution with shape $n$ and scale 1 $\lambda \sim \Gamma(n, 1)$. $\Gamma(n, 1)$ is the posterior distribution of $\lambda$ with a flat prior distribution. This allows us to integrate over uncertainty in the stomatal density from the sample image.

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/tessellation-example.pdf}
\caption{Examples of synthetic and real leaf surfaces.  A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface. The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
\label{fig:tessellation-example}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/density.pdf}
\caption{Stomatal density is higher in \textit{A. thaliana} plants grown under high light conditions. We determined statistical significance between light treatments using Tukey post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:density}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{figures/single-surface.pdf}
\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1). We determined statistical significance between light treatments using Tukey post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:single-surface}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces. Dashed line indicates zero correlation. Weak positive correlations are not significantly different from zero after correcting for multiple comparisons. The correlation does not differ among light treatments.}
\label{fig:dual-surface}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
\label{fig:length-area}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/model_summary.pdf}
\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis. The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane. `r symbol_descriptions`.}
\label{fig:model_summary}
\end{figure}

@@

Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017]. A theoretical, optimized plant would minimize stomatal density while also allowing competitive gas exchange rates for its environment so as to maximize C assimilation per unit investment in stomata. Natural selection operates within developmental and physical constraints to drive each plant species toward its theoretical optimum. This study provides evidence that stomata in *Arabidopsis thaliana*, the model angiosperm, are non-randomly distributed, favoring dispersion over clustering (\autoref{fig:single-surface}). However, stomata are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and standardize the area supplied by each stomate (\autoref{fig:tessellation-example}). Additionally, when grown in high light environments, *A. thaliana* exhibited increased stomatal density rather than increased stomatal dispersion (\autoref{fig:density}), which suggests that natural selection has acted more strongly on developmental pathways that modulate stomatal density than those that control stomatal dispersion. In other words, plants optimize gas exchange by adding more stomata rather than dispersing them more evenly across the leaf surface. This study also demonstrates that stomata that supply larger leaf areas with CO$_2$ tend to be larger (\autoref{fig:length-area}). These results could suggest that 1) the added energetic and hydraulic cost of non-ideally dispersed stomata is negligible and therefore not acted on by natural selection; 2) no developmental pathway exists to ensure the ideal placement of stomata on the leaf; or 3) the regulation of stomatal size limits the cost incurred by non-ideal stomatal dispersion.

In high light environments, amphistomy is favorable as high light photosynthesis is limited by CO$_2$ and amphistomy halves diffusion path length and boundary layer resistance, thereby reducing CO$_2$ limitation - increasing theoretical $A_\text{max}$. An optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}. However, our results show that leaf surfaces are not coordinated but are independent, regardless of light (\autoref{fig:dual-surface}). Additionally, gas exchange models show little photosynthetic efficiency gain from abaxial-adaxial stomatal coordination compared to anticoordination (INSERT FIG FROM MODELING). We posit that this marginal gain is not sufficient to be acted upon strongly by natural selection. Thus, amphistomatous plants do not exhibit abaxial-adaxial stomatal coordination for there is little selective advantage of it.

Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching the theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modelling should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

Beyond disperson on a single surface, gas exchange can be optimized via stomatal coordination of abaxial and adaxial surfaces in amphistomatous leaves. Given that leaf thicknesses are generally multiple times greater than interstomatal distance (GIVE DISTANCES HERE). As a result, abaxial-adaxial stomatal coordination reduces CO$_2$ diffusion path length far less than single surface dispersion, so we hypothesize this strategy to afford less photosynthetic advantage to the leaf. Our modelling results demonstrate that, even in ideal conditions, i.e. thick leaf, low stomatal densities, high light, low leaf porosity, high Rubisco concentration, etc., the photosynthetic advantage of coordination is minimal. We are not surprised by these results, but still highlight them here as we are the first to report this finding.

Amphistomy is a unique and important adaptation found around the world across many plant lineages [@muir_light_2018], yet much of the dynamics of amphistomy remain poorly understood. Here, we show that in *Arabidopsis thaliana* 1) stomata in are non-randomly dispersed, but not ideally dispersed; 2) stomatal size and density are modulated by light; 3) stomatal size is positively correlated with the area to which it supplies CO$_2$; and 4) abaxial-adaxial stomatal coordination is not exhibited and is not shown to provide a strong photosynthetic advantage using CO$_2$ diffusion models. Interestingly, these findings did not validate many of our hypotheses which were based on first principles, suggesting that there may be limits on plants' ability to control stomatal placement. Future studies which elucidate these limitations may have important implications for agriculturaly productivity in a rapidly changing world.

>>>>

---
title: Does stomatal patterning in amphistomatous leaves minimize the CO$_2$ diffusion path length with leaves of *Arabidopsis thaliana*?
date: "`r Sys.Date()`"
author:
  - name: Jacob L. Watts
    email: Jacob.Watts-1@colorado.edu
    corresponding_author: yes # Footnote using corresponding_author
    affiliation: 
      - UHM
      - CU
  - name: Graham Dow
    email: graham.dow@niab.com
    affiliation: NIAB
  - name: Thomas N. Buckley
    email: tnbuckley@ucdavis.edu
    affiliation: UCD
  - name: Christopher D. Muir
    email: cdmuir@wisc.edu
    affiliation: 
      - UHM
      - UWM
    # footnote: 1 # Footnote using reference
address:
  - code: UHM
    address: School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI 96822
  - code: CU
    address: Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO 80309
  - code: NIAB
    address: National Institute of Agricultural Botany, Cambridge, United Kingdom CB3 0LE
  - code: UCD
    address: Department of Plant Sciences, University of California, Davis, CA 95616
  - code: UWM
    address: Department of Botany, University of Wisconsin, Madison, WI 53706
# footnote:
#   - code: 1
#     text: "Current email address: \\href{mailto:cdmuir@wisc.edu}{cdmuir@wisc.edu}"
abstract:
  - "We will write abstract after Discussion and Conclusions are complete"
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
competing_interests: |
  The authors declare no competing interests.
author_contributions: |
  JLW and CDM conceived of the project, analyzed data, and wrote the manuscript. GW provided data. TNB contributed to model development and helped edit the manuscript.
acknowledgements: |
  This work was supported by NSF Award 1929167 to CDM and NSF Award 2307341 to TNB.
bibliography: stomata-spacing.bib
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
    oup_version: 1 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    # journal: "AoB PLANTS"
    document_style: "modern" # Can be contemporary, modern, traditional
    papersize: "large" # Can be large, medium, small
    #citation_package: "default" # Uncomment when using a CSL; default "natbib"
    namedate: TRUE # Set FALSE to use numeric refs; Default FALSE
    number_sections: FALSE # Uncomment to not number sections; default TRUE
    #number_lines: TRUE # Use `lineno` package - Default FALSE
    #number_lines_options: ["mathlines","switch"]  # Options for latex lineno package.
    onecolumn: TRUE # Uncomment for one column format; default FALSE
    extra_dependencies:
      - booktabs # to use with knitr::kable() example below
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}
 - \usepackage{caption} # for \captionsetup{}
 - \usepackage{newunicodechar,graphicx}
 - \DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
 - \newunicodechar{ʻ}{\okina}

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE,
  fig.pos = 'th',
  out.width = '100%', dpi = 300,
  fig.env = "figure"
)
knitr::opts_knit$set(root.dir = "..")
```

@@

Many successful predictions about stomata and other leaf traits can be made by hypothesizing that natural selection should optimizes CO$_2$ gain per unit of water loss [@cowan_stomatal_1977; @buckley_optimal_2017; @sperry_predicting_2017]. However, stomatal anatomy may be partially constrained by physical and developments limits on phenotypic expression [@croxdale_stomatal_2000; @harrison_influence_2020; @muir_how_2023]. Sometimes optimization leads to similar phenotypes across many disparate species. For example, almost all stomata follow the one cell spacing rule to maintain proper stomatal functioning [@geisler_oriented_2000; @dow_physiological_2014]; however some species (notably in *Begonia*) appear to benefit from overlapping vapor shells caused by stomatal clustering [@yi_gan_stomatal_2010; @lehmann_effects_2015; @papanatsiou_stomatal_2017]. Stomatal traits also vary adaptively in different environments. Stomatal density positively co-varies with irradiance during leaf development and negatively co-varies with CO$_2$ concentration [@gay_influence_1975; @schoch_dependence_1980; @woodward_stomatal_1987; @royer_stomatal_2001], consistent with optimality predictions. Stomatal size is jointly controlled by genome size, light, and stomatal density [@jordan_environmental_2015]. Size positively co-varies with genome size [@roddy_scaling_2020] and negatively co-varies with stomatal density [@camargo_density_2011]. Total stomatal area (size $\times$ density) is optimized for operational conductance ($g_\text{s,op}$) rather than maximum conductance ($g_\text{s,max}$) such that stomatal apertures are most responsive to changes in the environment at their operational aperture [@franks_physiological_2012; @liu_scaling_2021]. Stomatal aperture can compensate for maladaptive stomatal densities to an extent [@bussis_stomatal_2006], but stomatal density and size ultimately determine a leaf's theoretical $g_\text{s,max}$ [@sack_developmental_2016], which is proportional to $g_\text{s,op}$ [@murray_consistent_2020]. Additionally, low stomatal densities lead to irregular and insufficient CO$_2$ supply and reduced photosynthetic efficiency in areas far from stomata [@pieruschka_lateral_2006; @morison_lateral_2005], while high stomatal densities can reduce water use efficiency (WUE) [@bussis_stomatal_2006] and incur excessive metabolic costs [@deans_optimization_2020]. In most species, stomata occur on the abaxial (usually lower) leaf surface; but amphistomy, the occurrence of stomata on both abaxial and adaxial leaf surfaces, is also prevalent in high light environments with constant or intermittent access to sufficient water [@mott_adaptive_1982; @jordan_using_2014; @muir_light_2018; @drake_two_2019; @muir_is_2019]. Amphistomy effectively halves the CO$_2$ diffusion path length and boundary layer resistance by doubling boundary layer conductance [@parkhurst_adaptive_1978; @harrison_influence_2020; @mott_amphistomy_1991]. Historically, stomatal patterning in dicot angiosperms was thought to be random with an exclusionary distance surrounding each stomate [@sachs_developmental_1974]; however, the developmental controls of stomatal patterning are poorly understood and likely more complex than random development along the leaf surface. @croxdale_stomatal_2000] reviews three developmental theories which attempt to explain stomatal patterning in angiosperms: inhibition, cell lineage, and cell cycle, ultimately arguing for a cell cycle based control of stomatal patterning.

@@

We hypothesized that natural selection will favor stomatal patterning and distribution to minimize the diffusion path length. In amphistomatous leaves, this would be accomplished by 1) a dispersed, equilateral triangular distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface. Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light. Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest. We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$). This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.

\begin{figure}[ht]
\includegraphics[width = 4in]{figures/ideal-amphi-grid.pdf}
\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
\label{fig:ideal-amphi-grid}
\end{figure}

@@

df = tibble::tribble(
  ~trait, ~relationship,
  "leaf thickness", "+",
  "interstomatal distance", "+",
  "leaf porosity", "-", 
  # "lat.-vert. diffusion ratio", "-", 
  # "temperature", "-",
  # "pressure", "-",
  # "Rubisco concentration", "+",
  "light", "+"
)

@@

[Graham - can you fill in details on growth conditions, image aquisitiion, and image dimensions?]

@@

For each sample image, we also simulated $10^3$ synthetic data with $n$ stomata ideally dispersed in an equilateral triangular grid. For these grids, we integrated over plausible stomatal densities and then conditioned on stomatal grids with exactly $n$ stomata. The simulated stomatal count was drawn from a Poisson distribution with the mean parameter $\lambda$ drawn from a Gamma distribution with shape $n$ and scale 1 ($\lambda \sim \Gamma(n, 1)$). $\Gamma(n, 1)$ is the posterior distribution of $\lambda$ with a flat prior distribution. This allows us to integrate over uncertainty in the stomatal density from the sample image.

@@

\begin{figure}[ht]
\includegraphics[height = \textheight]{figures/tessellation-example.pdf}
\caption{Examples of synthetic and real leaf surfaces.  A) Uniform random synthetic leaf surface; B) Example of real leaf surface; C) Regularly distributed synthetic leaf surface. The zone defined by each stomate was calculated with voronoi tessellation and correlated with stomatal length in real leaves.}
\label{fig:tessellation-example}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = 4in]{figures/density.pdf}
\caption{Stomatal density is higher in \textit{A. thaliana} plants grown under high light conditions. We determined statistical significance between light treatments using Tukey post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:density}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=4in]{figures/single-surface.pdf}
\caption{Stomata are more dispersed than expected under the null model of uniform random position (dispersion index = 0) but far from a distribution that maximizes distance between stomata (dispersion index = 1). We determined statistical significance between light treatments using Tukey post-hoc tests. $^*~0.05 > P \ge 0.01; ^{**}~0.01 > P \ge 0.001; ^{***}~0.0001 > P \ge 0.0001; ^{***}~ P <0.0001$.}
\label{fig:single-surface}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = 4in]{figures/dual-surface.pdf}
\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces. Dashed line indicates zero correlation. Weak positive correlations are not significantly different from zero after correcting for multiple comparisons. The correlation does not differ among light treatments.}
\label{fig:dual-surface}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = 4in]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
\label{fig:length-area}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width = 5in]{figures/model_summary.pdf}
\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis. The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane. `r symbol_descriptions`.}
\label{fig:model_summary}
\end{figure}

@@

Stomata cost resources to maintain [@deans_optimization_2020] and expose leaves to risks such as hydraulic failure [@wang_theoretical_2020] or infection by plant pathogens [@melotto_stomatal_2017]. Therefore leaves should develop enough stomata to adequately supply CO$_2$ to chloroplasts, but not overinvest. A widespread hypothesis in plant ecophysiology is that natural selection optimizes traits like stomatal size, density, and distribution to maximize carbon gain relative to any costs in a given environmental context. In principle, spacing stomata to minimize the average distance between stomatal pores and chloroplasts within the mesophyll should increase carbon gain, all else being equal. However, reducing this distance to its absolute minimum may be constrained by developmental processes, or the photosynthetic benefit may be too small to be 'seen' by natural selection (i.e. the selection coefficient is less than drift barrier [@sung_drift-barrier_2012]).

We tested five related hypotheses about stomatal spacing in amphistomatous leaves using the model angiosperm *Arabidopsis thaliana* grown under different light intensities. First, we predicted that stomata on each surface are overdispersed relative to a random uniform distribution, which should increase CO$_2$ supply. Stomata on each surface are overdispersed (\autoref{fig:single-surface}), but are not ideally dispersed in an equilateral triangular grid as would be optimal to minimize CO$_2$ diffusion path length and equalize the area supplied by each stomate (\autoref{fig:tessellation-example}). Second, we predicted that an optimal amphistomatous leaf has offset stomata such that stomata are more likely to appear on one leaf surface if there is not a stomata directly opposite it on the other surface as shown in \autoref{fig:ideal-amphi-grid}. However, there is no evidence for coordination and the positions on each surface appear independent, regardless of light treatment (\autoref{fig:dual-surface}). Third, we predicted that plants respond plastically to higher light intensity by increasing stomatal density. *Arabidopsis* plants grown under high light had higher stomatal density than the same genotype growns under low and medium light intensity (\autoref{fig:density}). However, we found no support for our fourth prediction that stomata would be more evenly dispersed at high light intensity (\autoref{fig:single-surface}). Finally, we predicted that within leaf variation in stomatal size would correlate with stomatal spacing, as larger stomata can supply larger volumes of adjacent mesophyll. In all three light treatments, stomatal size positively covaried with the stomatal zone, i.e. adjacent region of mesophyll that would be supplied by that stomate (\autoref{fig:length-area}).

Stomatal spacing on *A. thaliana* leaves partially supports our overall hypothesis that natural selection minimizes the average distance between stomata and chloroplasts, for a given overall stomatal density. There are three nonmutually exclusive hypotheses for why several of our predictions were wrong. First, our predictions are wrong becuase they are based on overly simplistic assumptions about epidermal and mesophyll anatomy. Second, natural selection may be constrained by developmental processes that prevent phenotypes from reaching their adaptive optima. Third, the benefit of some traits may be of too little consequence to result in fitness differences large enough to be respond to selection. We consider the plausibility of these alternative hypotheses below and present ideas for future work to test them.

[this paragraph should discuss why our assumptions may be too simplistic]
We assume an idealized leaf epidermal and mesophyll structure that is homogenous and unconstrained by other tradeoffs. Real leaves not only provide pathways for CO$_2$ diffusion, but must supply water, intercept light, and deter herbivores and pathogens. These competing interests results in nonuniform epidermal and mesophyll structure that could alter our predictions about optimal stomatal spacing. [explain some examples]

[this paragraph should discuss developmental constraint] No developmental pathway exists to ensure the ideal placement of stomata on the leaf. [add more explanation]

[this paragraph should discuss limits on natural selection] The gas exchange model demonstrate little photosynthetic gain from abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}). Even though lateral diffusion may limit photosynthesis [@morison_lateral_2005], the marginal gain from optimally offsetting stomata is not sufficient to generate fitness differences relative to the strength of genetic drift (i.e. the drift-barrier). We can similarly extrapolate that an ideal, equilateral triangular stomatal spacing is only slightly better than a suboptimal overdispered pattern. Explaining these observations as the result of weak selection is in tension with the finding that stomatal size and zone positively covary, which would suggest that small changes in lateral diffusion distance are significant. As described above, the positive correlation between stomatal size and zone may be explained by common developmental processes rather than as an adaptation to maximize CO$_2$ diffusion.

[discuss significance of our study and future directions] Our study corroborates previous studies which demonstrate that stomata are non-randomly distributed along the leaf surface as a result of developmental mechanisms such as spatially biased arrest of stomatal initials [@boetsch_arrest_1995], oriented asymmetric cell division [@geisler_oriented_2000], and cell cycle controls [@croxdale_stomatal_2000]. We do not investigate the potential developmental pathways that influence stomatal dispersion in this study; however, they are important to consider as these pathways could limit plants from reaching the theoretical peak in the adaptive landscape: uniform stomatal dispersion. Instead, as this study suggests, plants may simply compensate with higher stomatal density and by fitting stomatal size to the area that they supply with CO$_2$. To understand why stomata are not ideally dispersed, more modelling should be done to estimate the fitness gain of stomatal dispersion. Additionally, genetic manipulation studies should attempt to create mutants with clustered and ideally dispersed stomata for a comparison of their photosynthetic traits. This could have extremely important implications for maximum assimilation rates in crops as most crop species are grown in high light where CO$_2$ is often limiting. In drought-prone environments, increased stomatal dispersion may increase water use efficiency by reducing the number of stomata needed to achieve the same internal CO$_2$ concentration, $C_\text{i}$.

[Conclusion] Our results suggest that after optimiziing stomatal density and having developmental rules for spacing stomata relatively evenly, there may be limited gains to further optimization. Therefore, developmental constraints may be necessary to make sense of some features of stomatal spacing and distribution.
